---
title: "Reproducible research project 1"
author: "Michael"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## import relevant R library
the following R library shall be imported for processing and analysis of data and setting working directory

```{r cache=TRUE}
setwd("./")
path<-"./"
if(!require("tidyverse")){
  install.packages("tidyverse")
  library(tidyverse)
}
if(!require("tidyr")){
  install.packages("tidyr")
  library(tidyr)
}

if(!require("dplyr")){
  install.packages("dplyr")
  library(dplyr)
}

if(!require("ggplot2")){
  install.packages("ggplot2")
  library(ggplot2)
}
```

## Importing data set

the dataset was downloaded from the url [download dataset][1],and imported using relevant R library

```{r cache = TRUE,results='hide'}

download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
              , destfile = paste(path, "dataFiles.zip", sep = "/"))
              unzip(zipfile = "dataFiles.zip")
              
data<-read.csv("./activity.csv")
head(data,n=6)
```
process dataset to remove NA's

```{r}
# data$day <- format(as.Date(data$date), "%d")               #stores the day from the date
data$day_of_week <- format(as.Date(data$date), "%A")       #stores the day of the week from the date
# str(data)
data <- data%>%
  mutate(day_period = case_when(day_of_week %in% c("Monday","Tuesday","Wednesday","Thursday","Friday") ~ "weekday",
                                TRUE ~ "Weekend"))  # Add a new variable called day_period to identify weekday or                                                         #weekend
factor_1<-factor(data$date)
```

create a summary dataset for sum of steps per day

```{r}
#data<-data%>%na_if("")%>%na.omit()
table_1<-data  %>%
  group_by(date) %>%
  summarise(sum_steps = sum(steps,na.rm = TRUE), Average_steps_per_interval = mean(steps,na.rm = TRUE),.groups = 'drop')

table_3<-data  %>%
  group_by(date) %>%
  summarise(median_steps_per_day = median(steps,na.rm = TRUE))

  
 head(table_1,n=20)
# data<-data%>%na_if("")%>%na.omit()
```


create a table showing average steps take on weekday and on weekends
```{r}
table_2<-data  %>%
  group_by(day_period) %>%
  summarise(Mean_day_peroid = mean(steps,na.rm = TRUE), .groups = 'drop')
table_2
```

## Including Plots

Histogram of the total number of steps taken each day

```{r pressure, echo=FALSE}
ggplot(table_1, aes(x = sum_steps)) +
    geom_histogram(fill = "orange") +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

# Plotting line graph for Time series plot of the average number of steps taken

```{r}
ggplot(table_1, aes(x = date , y = Average_steps_per_interval,group = 1 )) + geom_point(color="red", size=1) + geom_line(color="blue", size=1,group = 1 ) + labs(title = "Avg. Daily Steps", x = "Interval", y = "Avg. Steps per day")

```
Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
ggplot(table_2,aes(day_period,Mean_day_peroid)) +
  geom_bar(stat="identity",fill="green") +
  theme_bw()+
  labs(x="day period", y="Average steps per interval") + 
  labs(title="Average steps per interval")

```

The 5-minute interval that, on average, contains the maximum number of steps

```{r}
data[data$steps == max(data$steps), ]

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
[1]:https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip/ "download dataset"