---
title: "Reproducible research project 1"
author: "Michael"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## import relevant R library
the following R library shall be imported for processing and analysis of data and setting working directory

```{r}
setwd("./")
path<-"./"
if(!require("tidyverse")){
  install.packages("tidyverse")
  library(tidyverse)
}
if(!require("tidyr")){
  install.packages("tidyr")
  library(tidyr)
}

if(!require("dplyr")){
  install.packages("dplyr")
  library(dplyr)
}

if(!require("ggplot2")){
  install.packages("ggplot2")
  library(ggplot2)
}

if(!require("chron")){
  install.packages("chron")
  library(chron)
}

```

## Importing data set

the dataset was downloaded from the url [download dataset][1],and imported using relevant R library

```{r,results='hide'}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if (!(file.exists("./dataFiles.zip")))
{ download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
              , destfile = paste(path, "dataFiles.zip", sep = "/"))}
if (!file.exists("./activity.CSV"))
{unzip(zipfile  = "dataFiles.zip")}
              
data<-read.csv("./activity.csv")
head(data,n=6)
```
process dataset to remove NA's

```{r}
# data$day <- format(as.Date(data$date), "%d")               #stores the day from the date
data$day_of_week <- format(as.Date(data$date), "%A")       #stores the day of the week from the date
# str(data)
data <- data%>%
  mutate(day_period = case_when(day_of_week %in% c("Monday","Tuesday","Wednesday","Thursday","Friday") ~ "weekday",
                                TRUE ~ "weekend"))  # Add a new variable called day_period to identify weekday or                                                         #weekend
factor_1<-factor(data$date)
#convert time with chron package
data$time <- sprintf("%04d", data$interval)
#data$time
data$time <- paste0(substr(data$time, 1,2), ":", substr(data$time, 3,4), ":", "00")
data$time <- chron(times. = data$time, format = "h:m:s")
Sys.setenv(TZ='GMT')
factor_2<-factor(data$time)
```

create a summary dataset for sum of steps per day

```{r}
#data_no_na<-data%>%na_if("")%>%na.omit()
data_no_na <-data[complete.cases(data),]
table_1<-data_no_na  %>%
  group_by(date) %>%
  summarise(sum_steps = sum(steps,na.rm = TRUE), Average_steps_per_day = mean(steps,na.rm = TRUE),Daily_median = median(steps,na.rm = TRUE), .groups = 'drop')
```

sorting time series interval into weekdays or weekend

```{r}
# table_6 <- table_7
# 
# table_6$day_of_week <- format(as.Date(table_1$date), "%A")       #stores the day of the week from the date
# 
# table_6 <- table_6%>%
#   mutate(day_period = case_when(day_of_week %in% c("Monday","Tuesday","Wednesday","Thursday","Friday") ~ "weekday",
#                                 TRUE ~ "weekend"))


# data<-data%>%na_if("")%>%na.omit()
```
create a dataset of average by time interval
```{r}

table_7<-data_no_na  %>%
  group_by(time) %>%
  summarise(sum_steps = sum(steps,na.rm = TRUE), Average_steps_per_interval = mean(steps,na.rm = TRUE), .groups = 'drop')


```
create a table showing average steps take on weekday and on weekends
```{r}
table_2<-data  %>%
  group_by(day_period) %>%
  summarise(Mean_day_peroid = mean(steps,na.rm = TRUE), .groups = 'drop')

mean_weekday <- table_2[table_2$day_period == "weekday","Mean_day_peroid"][[1]]
mean_weekend <- table_2[table_2$day_period == "weekend","Mean_day_peroid"][[1]]
```

 the strategy used for imputing missing data shall consist of replacing na value with evaluated mean for weekdays and weekend as appropriate

```{r}
data_na_imputted <- data #create a copy of the dataset
weekday_na <- is.na(data_na_imputted$steps) & data_na_imputted$day_period == "weekday"
weekend_na <- is.na(data_na_imputted$steps) & data_na_imputted$day_period == "weekend"
data_na_imputted[weekday_na,"steps"] <- mean_weekday
data_na_imputted[weekend_na,"steps"] <- mean_weekend

table_5<-data_na_imputted  %>%
  group_by(date) %>%
  summarise(sum_steps = sum(steps,na.rm = TRUE), Average_steps_per_interval = mean(steps,na.rm = TRUE),.groups = 'drop')
```

sorting different dataset for weekday and weekend

```{r}
data_weekdays <- data[data$day_period == "weekday",]
data_weekends <- data[data$day_period == "weekend",]
table_3<-data_weekdays  %>%
  group_by(time) %>%
  summarise(sum_steps = sum(steps,na.rm = TRUE), Average_steps_per_interval = mean(steps,na.rm = TRUE),.groups = 'drop')
table_4<-data_weekends  %>%
  group_by(time) %>%
  summarise(sum_steps = sum(steps,na.rm = TRUE), Average_steps_per_interval = mean(steps,na.rm = TRUE),.groups = 'drop')

```


## Including Plots

Histogram of the total number of steps taken each day

```{r pressure, echo=FALSE}
ggplot(table_1, aes(x = sum_steps)) +
    geom_histogram(fill = "orange",binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```
# Histogram with na values imputted
```{r}
ggplot(table_5, aes(x = sum_steps)) +
    geom_histogram(fill = "orange") +
    labs(title = "Daily Steps NA imputted", x = "Steps", y = "Frequency")
```

# Plotting line graph for Time series plot of the average number of steps taken

```{r}
ggplot(table_7, aes(x = time , y = Average_steps_per_interval,group = 1 )) + geom_point(color="red", size=1) + geom_line(color="blue", size=1,group = 1 ) + labs(title = "Avg. Daily Steps", x = "Interval", y = "Avg. Steps per day")

```
Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
ggplot(table_2,aes(day_period,Mean_day_peroid)) +
  geom_bar(stat="identity",fill="green") +
  theme_bw()+
  labs(x="day period", y="Average steps per interval") + 
  labs(title="Average steps per interval")

```
#plotting average daily steps on weekday and weekends
```{r}
ggplot(table_3, aes(x = time , y = Average_steps_per_interval,group = 1 )) + geom_point(color="red", size=1) + geom_line(color="red", size=1,group = 1 ) + labs(title = "Avg. Daily Steps weekdays", x = "Days", y = "Avg. Steps per day weekdays")

ggplot(table_4, aes(x = time , y = Average_steps_per_interval,group = 1 )) + geom_point(color="blue", size=1) + geom_line(color="blue", size=1,group = 1 ) + labs(title = "Avg. Daily Steps weekends", x = "Interval", y = "Avg. Steps per day")
```


The 5-minute interval that, on average, contains the maximum number of steps

```{r}
table_7[table_7$Average_steps_per_interval == max(table_7$Average_steps_per_interval), ]

# #remove missing values from vector
# x <- x[complete.cases(x)]
# 
# #remove rows with missing values in any column of data frame
# df <- df[complete.cases(df), ]
# 
# #remove rows with NA in specific columns of data frame
# df <- df[complete.cases(df[ , c('col1', 'col2', ...)]), ] 
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
[1]:https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip/ "download dataset"